# sleepstudy dataset

Lets now consider the sleepstudy dataset from the lme4 package. This is from a
report on a study of the effects of sleep deprivation on reaction time for a
number of subjects chosen from a population of long- distance truck drivers.
These subjects were divided into groups that were allowed only a limited amount
of sleep each night. We consider here the group of 18 subjects who were
restricted to three hours of sleep per night for the first ten days of the
trial. Each subject’s reaction time was measured several times on each day of
the trial.

```{r}
library(lme4)
str(sleepstudy)
head(sleepstudy)
```

As always, we start by plotting the data

```{r}
library(ggplot2)
qplot(Days, Reaction, facets=~Subject, data = sleepstudy)
```

qplot has an option to also draw on a fitted regression line to each person as
well.

```{r}
c <- ggplot(sleepstudy, aes(Days, Reaction))+geom_point() + facet_wrap(~ Subject, nc=5)
c+stat_smooth(method='lm')
```

## Model 1: Correlated random effects

```{r}
(fm06 <- lmer(Reaction ~ 1 + Days + (1 +Days | Subject), sleepstudy))
```

+ What is the typical initial reaction time?
+ How much does reaction time typically increase per day?
+ What is typical subject-subject variation in the initial reaction time?
+ What is the typical subject-subject variation in the slope?
+ What would approximate 95% confidence intervals be for the slope and the
intercept across all subjects?
+ What is the typical within subject variation?
+ Is there a strong relationship between a subjects initial reaction time and
how strongly affected they are by sleep deprivation?

This model has used correlated random effects for the same subject, i.e., there
is a correlation between the random subject intercept and the subject specific
gradient. Mathematically, we have fit the model

$$ y_{ij} = \alpha + a_i + (\beta+ b_i)\mbox{Days}_j + \epsilon_{ij}$$

where
$$ \left(\begin{array}{c}a_i\\b_i\end{array}\right)\sim N(0, \Sigma) \mbox{ and
} \epsilon_{ij} \sim N(0, \sigma^2)$$

The output above gives us estimates of the fixed effects $(\alpha, \beta)$, and
the random effect variances $\Sigma$ and $\sigma^2$.





















## Model 2: Uncorrelated random effects

```{r}
(fm07 <- lmer(Reaction ~ 1 + (1 | Subject)+ Days + (Days-1|Subject), sleepstudy))
```

Note that we have explicitly ruled out the intercept term in the random effect
part for Days. If we didn't do this, R would have fit an additional
(unnecessary) intercept.

Note that this has fit the model

$$ y_{ij} = \alpha + a_i + (\beta+ b_i)\mbox{Days}_j + \epsilon_{ij}$$

where
$$ a_i\sim N(0, \sigma^2_a), \quad  b_i\sim N(0, \sigma^2_b) \mbox{ and }
\epsilon_{ij} \sim N(0, \sigma^2)$$
with all the random effects independent of each other.






In the correlated random effects model, the correlation was estimated to be
small. This suggests that the uncorrelated random effects model is a good
choice. We will look later at how to formally test which model is better.
